---
title: "Multivariate-EMS-Code"
author: "Sanjana Bajaj"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

# check which columns have NAs or empty cells
sapply(emergency_response_data, function(col) {
  sum(is.na(col) | (is.character(col) & col == ""))
})

```

```{r}

# drop columns that are redundant or irrelevant for response time
emergency_response_data <- emergency_response_data %>%
  select(
    -Box,
    -Address,
    -`Unit ID`,
    -`Call Number`,
    -`Original Priority`,
    -`Zipcode of Incident`,
    -`Available DtTm`,
    -`Transport DtTm`,
    -`Hospital DtTm`,
    -`Watch Date`,
    -`Response DtTm`,
    -`Supervisor District`,
    -`Unit sequence in call dispatch`,
    -`Final Priority`,
    -`Station Area`,
    -`Battalion`,
    -`Call Type Group`,
  )

```

```{r}
emergency_response_data <- emergency_response_data %>% select(-`Entry DtTm`)

```

```{r}
emergency_response_data[emergency_response_data == ""] <- NA

# drop all the rows with NAs/empty cells for necessary features
emergency_response_data <- emergency_response_data %>%
  filter(
    !is.na(case_location),                     
    !is.na(`On Scene DtTm`),
    !is.na(City)
  )
```

```{r}
# some SF neighborhoods were marked in the "city" col. these should say SF since they are part of SF

emergency_response_data$City[emergency_response_data$City %in% c("Presidio", "Treasure Isla", "Hunters Point", "Fort Mason", "Yerba Buena")] <- "San Francisco"


```

```{r}

sapply(emergency_response_data, function(col) {
  sum(is.na(col) | (is.character(col) & col == ""))
})

```

```{r}
library(lubridate)
library(dplyr)

# convert to date-time
emergency_response_data <- emergency_response_data %>%
  mutate(
    `Received DtTm`  = ymd_hms(`Received DtTm`),
    `On Scene DtTm`  = ymd_hms(`On Scene DtTm`),
    `Dispatch DtTm` = ymd_hms(`Dispatch DtTm`)
  )

# create the response_time variable
emergency_response_data <- emergency_response_data %>%
  mutate(
    response_time = as.numeric(difftime(`On Scene DtTm`, `Received DtTm`, units = "mins"))  )

```

```{r}

# remove response & dispatch times < 0, and the one incident with 1911 min response & dispatch time

emergency_response_data <- emergency_response_data %>%
  filter(response_time > 0)


```

```{r}
head(sort(emergency_response_data$response_time, decreasing = TRUE), 30)

# drop the rows with response time > 180 mins -- this is very likely to be a special case, or a data entry error

emergency_response_data <- emergency_response_data %>%
  filter(response_time <= 180)

```

```{r}


emergency_response_data <- emergency_response_data %>%
  mutate(
    hour = hour(`Received DtTm`),
    day_of_week = wday(`Received DtTm`, label = TRUE),
    month = month(`Received DtTm`, label = TRUE),
    year = year(`Received DtTm`)
  )

```

```{r}
emergency_response_data <- emergency_response_data %>% select(-`Call Date`)

```

```{r}
emergency_response_data <- emergency_response_data %>% select(-`Unit Type`)


```


```{r}
## EXPLORATORY DATA ANALYSIS

summary(emergency_response_data)

```

```{r}

hist(emergency_response_data$response_time)
hist(log1p(emergency_response_data$response_time))


```


```{r}
# visualize response time by hour of day, day of week, month

# hour of day (y range is small (like 2 mins), but there is a spike in the daytime)
emergency_response_data %>%
  group_by(hour) %>%
  summarise(mean_rt = mean(response_time)) %>%
  plot()

```

```{r}

# day of week (weekdays higher, just by a bit)
emergency_response_data %>%
  group_by(day_of_week) %>%
  summarise(mean_rt = mean(response_time)) %>%
  plot()

```

```{r}

# month (even tighter y range, kind of all over the place)
emergency_response_data %>%
  group_by(month) %>%
  summarise(mean_rt = mean(response_time)) %>%
  plot()

```

```{r}

# response time by call type
emergency_response_data %>%
  group_by(`Call Type`) %>%
  summarise(mean_rt = mean(response_time), n = n()) %>%
  arrange(desc(mean_rt))

```

```{r}
# priority
emergency_response_data %>%
  group_by(Priority) %>%
  summarise(mean_rt = mean(response_time), n=n())
```

```{r}
# neighborhood
emergency_response_data %>%
  group_by(`Neighborhooods - Analysis Boundaries`) %>%
  summarise(mean_rt = mean(response_time), n=n()) %>%
  arrange(desc(mean_rt))

# so 10 incidents have "None"
# western addition is fastest - google confirms - has a lot of private providers
# presidio heights, pac heights, rich asf
```

```{r, fig.height=10}
boxplot(response_time ~ `Fire Prevention District`, data = emergency_response_data, outcex=0.5)

```

```{r, fig.height=10, fig.width=8}
boxplot(response_time ~ fire_response, data = emergency_response_data, outcex=0.3)

```

```{r, fig.height=10, fig.width=8}
boxplot(response_time ~ medical_als, data = emergency_response_data, outcex=0.3)

```


```{r}
# most common incidents by neighborhood
neighborhood_calls <- emergency_response_data %>%
  group_by(`Neighborhooods - Analysis Boundaries`, `Call Type`) %>%
  summarise(n = n(), .groups = "drop")

top_calls <- neighborhood_calls %>%
  group_by(`Neighborhooods - Analysis Boundaries`) %>%
  slice_max(order_by = n, n=10)

top_calls

```

```{r, fig.width = 13}

# visualize
library(ggplot2)

ggplot(top_calls, aes(x = reorder(`Neighborhooods - Analysis Boundaries`, n), y = n, fill = `Call Type`)) +
  geom_col() +
  coord_flip()

```

```{r}
## K MEANS

# no grouping
call_data <- emergency_response_data %>%
  select(
    response_time,
    medical_als,
    medical_bls,
    fire_response,
    command,
    support,
    hour,
    Priority,
    `Number of Alarms`,
    day_of_week
  )

# don't include neighborhood + call type -- too many dummy vars would be created
# see afterwards - which neighborhoods/call types dominate each cluster

library(fastDummies)
call_data_encoded <- fastDummies::dummy_cols(
  call_data,
  select_columns = c("day_of_week", "Priority"),
  remove_first_dummy = TRUE,   # avoids multicollinearity
  remove_selected_columns = TRUE
)

```


```{r}
# scale
X1 <- scale(call_data_encoded)

```

```{r}
# pick k
elbow <- sapply(1:10, function(k) {
  kmeans(X1, centers=k, nstart=10)$tot.withinss
})
plot(1:10, elbow,type="b")

#5?

```



```{r}
set.seed(21)

k1 <- kmeans(X1, centers = 5, nstart = 25)

call_data_encoded$cluster <- k1$cluster

```

```{r}
call_data_encoded %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean))

```

```{r}
library(ggplot2)

ggplot(call_data_encoded, aes(x = factor(cluster), y = response_time)) +
  geom_boxplot() +
  labs(x = "Cluster", y = "Response Time")

```

```{r}
# leave only rows with priority=2 (non emergency), and priority=3 (emergency), which is what the documentation describes (other character codes are less frequent anyways.. except 1, but we have no way of knowing for sure what it means).

# remove hour, number of alarms, day of week. these vars had basically the same value across clusters.

emergency_response_data %>%
  count(Priority, sort = TRUE)

call_data_2 <- emergency_response_data %>%
  select(
    response_time,
    medical_als,
    medical_bls,
    fire_response,
    command,
    support,
    Priority
  )

call_data_2 <- call_data_2 %>%
  filter(Priority %in% c("2", "3")) %>%
  mutate(Priority = as.numeric(Priority))

```


```{r}
# scale
X2 <- scale(call_data_2)

```

```{r}
# pick k
elbow <- sapply(1:10, function(k) {
  kmeans(X2, centers=k, nstart=10, iter.max=50)$tot.withinss
})
plot(1:10, elbow,type="b")

#4?

```

```{r}

set.seed(21)

k2 <- kmeans(X2, centers = 4, nstart = 25)

call_data_2$cluster <- k2$cluster

```

```{r}
call_data_2 %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean))

```
```{r}
ems_filtered <- emergency_response_data %>%
  filter(Priority %in% c("2", "3")) %>%
  mutate(Priority = as.numeric(Priority))

nrow(ems_filtered)
nrow(call_data_2)

```

```{r}
ems_filtered$cluster2 <- k2$cluster # cluster2 = kmeans attempt 2 clustering

```

```{r, fig.width=9}
# what is each cluster like
call_data_2 %>%
  group_by(cluster) %>%
  summarise(
    rt = mean(response_time),
    als = mean(medical_als),
    bls = mean(medical_bls),
    fire = mean(fire_response),
    command = mean(command),
    support = mean(support),
    priority = mean(Priority)
  ) %>%
  tidyr::pivot_longer(-cluster) %>%
  ggplot(aes(x = name, y = value, fill =factor(cluster))) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

- Cluster 1 (Red) [Low-priority, slow BLS/service calls]: No/low ALS, BLS-heavy, low command, some fire, non-emergency, high response time, no support

- Cluster 2 (Green) [High-priority, fast ALS (+support?) calls]: ALS-heavy, some BLS, low command, low fire, emergency, low response time, SUPPORT
  - extra support could mean potentially more complicated/serious scenes

- Cluster 3 (Blue) [Non-emergency or delayed ALS calls]: ALS-heavy, no/low BLS, no/low command, no/low fire, non-emergency, highest response time, no support

- Cluster 4 (Purple) [Fire Incidents]: ALS-heavy, low BLS, low command, FIRE, emergency, LOWEST response time, no support

```{r}
# confirm/understand further by looking at call type distribution by cluster
calltype_cluster <- ems_filtered %>%
  count(cluster2, `Call Type`) %>%
  group_by(cluster2) %>%
  mutate(prop = n / sum(n))

calltype_cluster %>%
  group_by(cluster2) %>%
  slice_max(prop, n = 10) %>%
  ggplot(aes(x = reorder(`Call Type`, prop),y = prop,fill = factor(cluster2))) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Call Types per Cluster", fill = "Cluster")

```

```{r, fig.width=10}
# remove "medical incident" bc it dominates
calltype_cluster <- ems_filtered %>%
  filter(`Call Type` != "Medical Incident") %>%
  count(cluster2, `Call Type`) %>%
  group_by(cluster2) %>%
  mutate(prop = n / sum(n))

calltype_cluster %>%
  group_by(cluster2) %>%
  slice_max(prop, n = 10) %>%
  ggplot(aes(x = reorder(`Call Type`, prop),y = prop,
             fill = factor(cluster2))) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Call Types per Cluster (Exclude Medical Incident)",
       fill = "Cluster")

```

- this explains blue - the ALS/emergency group with long response times - citizen assistance/service call is complicated

numerically cluster 4 dominates (fires):
1        4 230916
2        3  96565
3        1  67116
4        2  15784

```{r}
# top neighborhoods per cluster
neighborhood_cluster <- ems_filtered %>%
  count(cluster2, `Neighborhooods - Analysis Boundaries`) %>%
  group_by(cluster2) %>%
  mutate(prop = n / sum(n))


neighborhood_cluster %>%
  group_by(cluster2) %>%
  slice_max(prop, n = 10) %>%
  ggplot(aes(x = reorder(`Neighborhooods - Analysis Boundaries`, prop),
             y = prop,
             fill = factor(cluster2))) +
  geom_col() +
  coord_flip() +
  labs(title = "Top Neighborhoods per Cluster", fill = "Cluster")
```

```{r, fig.width=10, fig.height=12}
ems_filtered %>%
  count(`Neighborhooods - Analysis Boundaries`, cluster2) %>%
  group_by(`Neighborhooods - Analysis Boundaries`) %>%
  mutate(prop = n / sum(n)) %>%
  ggplot(aes(x = reorder(`Neighborhooods - Analysis Boundaries`, prop),y = prop,fill = factor(cluster2))) +
  geom_col(position = "fill") +
  coord_flip()
```

```{r}
# PCA

pca <- prcomp(X2)

pca_df <- data.frame(pca$x[,1:2], cluster = factor(k2$cluster))

ggplot(pca_df, aes(PC1, PC2, color = cluster)) +
  geom_point(alpha = 0.3)

```

```{r}
# remove response time?

call_data_3 <- emergency_response_data %>%
  select(
    medical_als,
    medical_bls,
    fire_response,
    command,
    support,
    Priority
  )

call_data_3 <- call_data_3 %>%
  filter(Priority %in% c("2", "3")) %>%
  mutate(Priority = as.numeric(Priority))

```

```{r}
# scale
X3 <- scale(call_data_3)

set.seed(21)

k3 <- kmeans(X3, centers = 4, nstart = 25)

call_data_3$cluster <- k3$cluster

```

```{r}
call_data_3 %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean))

```

```{r, fig.height=12, fig.width=8}
ems_filtered$cluster3 <- k3$cluster

ggplot(ems_filtered, aes(x = factor(cluster3), y = response_time)) +
  geom_boxplot() +
  labs(
    x = "Cluster",
    y = "Response Time (minutes)",
    title = "Response Time Distribution Across Clusters (No RT in Clustering)"
  )

```

```{r, fig.width=9}
# what is each cluster like
call_data_3 %>%
  group_by(cluster) %>%
  summarise(
    als = mean(medical_als),
    bls = mean(medical_bls),
    fire = mean(fire_response),
    command = mean(command),
    support = mean(support),
    priority = mean(Priority)
  ) %>%
  tidyr::pivot_longer(-cluster) %>%
  ggplot(aes(x = name, y = value, fill =factor(cluster))) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
- cluster numbers shuffled, but basic meaning of categories appears the same -- double check this


```{r}




```

```{r}


```




